Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-12 ThesisJP/Introduction/introduction.tex
Started on Introduction 2.5.

\section{Discussion and future work}
Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the maximum throughput of the target GPU. This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended. There might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed. This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU. An interesting observation can be found in Fig.\,10 of \textbf{Paper\,I}. Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}. This clearly shows the benefit of modern GPU computing.  Solving was performed by a batched Gauss Jordan solver implemented by Nvidia. In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home. Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible. Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might also help to improve the solver's throughput.
\\\\
In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. %Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally. 
The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}. 
\\\\
In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments. Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on polar-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 
As just mentioned in the discussion of \textbf{Paper\,II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS. It will, however,  not correct for signal cancellation caused by phase aberrations. At best it will correct for the linear term introduced by the aberrator if the term is small.
With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided and the shift-invariance property is maintained. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.
\\\\
%Objections to \textbf{Paper\,IV} includes its inherent chicken-or-egg dilemma. The visualization is adjusted based on the data acquired by a user. The user sees the visualization and adjust scanning based on the visual input. This recursive behavior suggest that the method is best suited for post-processing when data has been recorded and saved. 
%Does data need to be saved in order to perform tracking?
One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Fig.\,3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed. A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function. A local opacity transfer function is just to extreme to be clinically robust. In addition, it would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less smoothing and higher opacity at tissue interfaces. If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.
\\\\
The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on the GPU. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could then also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.  In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible. As of January 2014 the simulator has been download 200 times and counting. It also has close to a "whopping" 1000 views on youtube. 
\\\\
Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been made open source and is available on github. 
\endinput

>>>>

\section{Discussion and future work}
\subsection{Paper\,I}
Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the peak theoretic throughput of the target GPU. This shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended. There might be ways to improve on these numbers, but both the matrix equation solver and the covariance matrix estimation step consist of several points where fine granular synchronization and serialized instructions are needed. These points will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU. The work in \textbf{Paper I} also showed that the covariance estimation kernel was memory bound, hence there is not enough instructions to hide all the memory latency. Further improvements through improved memory management could therefore be possible.

An interesting observation in \textbf{Paper\,I} can be found in Fig.\,10 \todo{Direct ref}. Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a high-end \nom{CPU}{Central processing unit} (The target CPU is from 2010). Note that the target GPU, the Nvidia Quadro 6000, is also from 2010. This clearly shows the benefit of modern GPU computing.

The matrix equation found in the Capon beamformer is solved using a batched Gauss Jordan solver implemented by Nvidia. In our work we focused on the covariance matrix estimation step, and did not try to beat Nvidia at home. Nevertheless, \textbf{Paper\,I} showed that the solver had significantly lower throughput than the covariance matrix estimation kernel. A natural next step would therefore be to analyze the current solver, and figure out if any further optimization is possible. Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition should also help to improve the throughput.

\subsection{Paper\,II}
In \textbf{Paper\,II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper\,II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper\,III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper\,II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper\,II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was mainly selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. %Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally. 
The work in \textbf{Paper\,II} made it clear that interesting findings are done when users are able to watch the temporal behavior of an algorithm. This led us to the work of \textbf{Paper\,III}.

\subsection{Paper\,III}
In \textbf{Paper\,III} we investigate the inherent self nulling involved with Capon beamforming, and how this effect comes naturally into play when the beamformer is subjected to linear steering vector errors caused by lateral undersampling. These errors are naturally reduced when the lateral sampling is increased. This simple observation is something which has been missing in the literature on Capon beamforming for medical ultrasound imaging. The reason could be ignorance or its possible negative impact on imaging frame rate. A lack of studies where Capon beamforming is performed on consecutive frames might also explain the lack of comments. Fortunately, as shown in \textbf{Paper\,III}, it is possible to maintain the same frame rate as with DAS beamforming by applying oversampling on receive using a set of steering vectors (phase delays). It also turns out that more matrix inversions are not needed either. Oversampling on receive will, however,  increase the penalty of applying post-processing, like filtering,  on polar-grid data. %The time taken to convert data from a polar scan grid to a Cartesian display grid will also increase. The Cartesian grid should also have high resolution in order to preserve the un-sharpening introduced by Capon beamforming. 
As just mentioned in the discussion of \textbf{Paper\,II}, reduced signal cancellation should prevent contrast from degrade to less than the contrast obtain with DAS. It will, however,  not correct for signal cancellation caused by phase aberrations. At best it will correct for the linear term introduced by the aberrator if the term is small.
With improved shift-invariance it will also be interesting to see if high resolution beamforming could act as a post-processing step for other algorithms. For example to increase the precision of speckle and edge tracking. In the case where the image is generated for post processing it does not need to be visually pleasant anymore, as long as signal cancellation is avoided and the shift-invariance property is maintained. Good looking speckle, or DAS-like speckle, has been the driving force behind many of the robustification techniques and the default parameters seen in the literature. Hence, more aggressive parameters might be used if the image produced is intended as input to another algorithm and not for the human eye.

\subsection{Paper\,IV}
%Objections to \textbf{Paper\,IV} includes its inherent chicken-or-egg dilemma. The visualization is adjusted based on the data acquired by a user. The user sees the visualization and adjust scanning based on the visual input. This recursive behavior suggest that the method is best suited for post-processing when data has been recorded and saved. 
%Does data need to be saved in order to perform tracking?
One limitation of the algorithm described in \textbf{Paper\,IV} is its dependence on a successful tracking. If the tracking fails the proposed method will fail.  Another issue is the view-dependent visibility, where an object could become visible or disappear while rotating or if the local statistics change a lot during the cardiac cycle. Even if the method successfully removes noise, this means that the rendering will become non-intuitive to interact with and unrealistic to watch. However, the method could still be used to clean up still images of standard cardiac views. The paper does also propose an interesting method for automatic tuning of rendering parameters. Fig.\,3d in \textbf{Paper\,IV} actually shows that an optimal threshold could be found based on global and not only local statistics. Even though the visualization based on global statistics has more errors, with respect to endocardium visibility, the tuning of the opacity function is still fully automatic if an iterative search for the minima in Fig.\,3d in \textbf{Paper\,IV} is performed. A global opacity transfer function will assure a rendering which are realistic, but it will still be dependent on a successful tracking for the auto-adjustment to work. An automatically tuned opacity function based on global statistics combined with an improved smoothing scheme is likely to be better suited for cardiac ultrasound imaging than a local opacity transfer function. A local opacity transfer function is just to extreme to be clinically robust. In addition, it would also be interesting to see if statistics based on the window selected by the Capon or the LCA beamformer could be utilized to adjust the opacity. It should also be possible to use this statistic as input to image filtering. An adaptively selected window will be symmetric in homogeneous regions and left-right shifted close to tissue interfaces. In that way we can apply less smoothing and higher opacity at tissue interfaces. If this will be different than just looking at neighboring pixels in an edge-preserving image filter is to be seen.

\subsection{Paper\,V}
The fifth paper stands out from the other papers by not dealing with an adaptive processing technique. It is still, however, centered around the topic of implementing algorithms for ultrasound imaging on the GPU. Ultrasound simulations are typically used for research and sometimes also to pre-calculate configurations of a given scan sequence, for instance to achieve a certain pressure in a given area. The continuously growth in computational power will, in the future, make it possible to evaluate a scan sequence on the fly, removing the need for lookup tables. Estimation of the mechanical and thermal index could then also be improved. Achieving higher processing speed for an excising algorithm by implementing it on a new architecture is not academically relevant in its own. It has to be combined with either a good analysis of what special adjustments that had to be made to the algorithm in order for the increase to happen or a suggestion of how to utilize the speed gained. These are points we have tried to answer in both \textbf{Paper\,I}, \textbf{II},  \textbf{IV}, and \textbf{V}.  In \textbf{Paper\,V} the increased speed made an interactive ''Paint-program'' feasible. As of January 2014 the simulator has been download 200 times and counting. It also has close to a "whopping" 1000 views on youtube. 
\\\\\\
Finally it is worth mentioning that the code for both the Capon beamformer\footnote{github.com/jpaasen/cos} and the ultrasound field simulator\footnote{github.com/jpaasen/hos} have been made open source and is available on github. 
\endinput
